Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-09-10 pnas/Towards.tex
Revised files. Will be sent to Amit for a second round (and maybe to Fred)

\begin{figure*}[h]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of micrographs of size $250\times 250$ with additive white Gaussian noise of variance $\sigma^2$ for increasing values of $\sigma$. Each micrograph contains the same four occurrences of a $50 \times 50$ image of Einstein. In panel (c), the noise level is such that it is very challenging to locate the occurrences of the planted image. In fact, it can be shown that at low SNR, reliable detection of individual image occurrences is impossible, even if the true image is known. By analogy to cryo-EM, this depicts a scenario where particle picking cannot be done.}
	 \vspace{-14pt} 	
\end{figure*}

\begin{figure*}[h]
	\centering
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_1}
		\caption{$500$ micrographs}
	\end{subfigure} 
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_4}
		\caption{$2\times 10^3$ micrographs}
	\end{subfigure} 
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_40}
		\caption{$2\times 10^4$ micrographs}
	\end{subfigure} 
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_400}
		\caption{$2\times 10^5$ micrographs}
	\end{subfigure}
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized randomly. Estimates are obtained from $500,2\times 10^3,2\times 10^4,2\times 10^5$ micrographs (growing across panels), each containing $700$ image occurrences on average.  \TODO{Tamir: I am redoing the exp}}	
 \vspace{-14pt} 
\end{figure*}

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical SNR. This fact has been recognized early on by the cryo-EM community. In particular, in an influential paper from 1995, Richard Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{myquote}{0in}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{myquote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight (below $\sim$50 kDa). 
Joachim Frank voices a similar observation in his 2017 Nobel prize lecture: ``\emph{Using the ribosome as an example, it became clear from the formula we obtained that the single-particle approach to structure research was indeed feasible for molecules of sufficient size: Particle Size > 3/[Contrast$^2$ $\times$ Resolution (as length) $\times$ Critical Electron Dose]}''~\cite{frank2018single}. 
As these two leaders of the cryo-EM community point out, it is impossible to reconstruct small molecules by any of the existing computational pipelines for single particle analysis in cryo--EM, because the particles themselves cannot be picked from the micrographs. Yet, imaging such molecules is important for a number of applications~\cite{scapin2018cryo}

@@

We observe $X$, and our task is to estimate $\eta$. How reliably can this be done? If $q \geq 1/2$, the constant estimator $\hat{\eta} = 0$ is correct with probability $q$; likewise, if $q \leq 1/2$, the constant estimator $\hat{\eta} = 1$ is correct with probability $1-q$. The question is, can we do better than this?  We prove that, as $\sigma \to \infty$, the answer is no:
\begin{proposition} \label{prop:two_gauss}
	For any deterministic estimator $\hat{\eta }$ of $\eta$,
	\begin{align*}
	\lim_{\sigma \to \infty} \P[ \hat{\eta} = \eta  ] & \leq \max(q, 1-q);
	\end{align*}
	that is: as the SNR deteriorates, the probability of success is no better than random chance.
\end{proposition}
\noindent The result is proved in Appendix~\ref{sec:proof_two_gauss}.

@@

From Propositions~\ref{prop:uniqueness} and~\ref{prop:gamma_sigma} we   deduce the following:
\begin{corollary}
	%Let $K=1$ and $\sigma > 0$ be fixed.
	In the limit of $N\to \infty$ and under the separation condition~\eqref{eq:spacing}, the signal $x$, the ratio $\gamma$, and the noise level $\sigma$ are determined from the first three autocorrelation functions of $y$ if either the signal $x$ is generic or $x$ has nonzero mean  and $\gamma > \frac{1}{4}$.
\end{corollary}

%This can be added to an arxiv version of the paper	
%From Propositions~\ref{prop:uniqueness} and~\ref{prop:gamma_sigma} we   deduce the following:
%\begin{corollary}
%	%Let $K=1$ and $\sigma > 0$ be fixed.
%	 In the limit of $N\to \infty$ and under the separation condition~\eqref{eq:spacing}, the signal $x$, the ratio $\gamma$, and the noise level $\sigma$ are determined from the first three autocorrelation functions of $y$ if:
%	\begin{itemize}
%		\item Either the signal $x$ is generic; or
%		\item   $x$ has nonzero mean  and $\gamma > \frac{1}{4}$.
%	\end{itemize}
%\end{corollary}
%
%\TODO{Similar results hold for signals in higher dimensions.---need to keep this sentence in red until the very end to make sure it's correct.}

@@

\begin{figure}[t]
	\centering
	\includegraphics[scale=0.28]{Einstein_recovery_error_combined}
	\caption{\label{fig:error_per_micro}Relative error curves for the experiment of Figure~\ref{fig:Einst_example}.}
\end{figure}

@@

{\bf Case 4:} $0=\ell_2 = \ell_1$. The same argument as for term (4) in Case 2 shows that terms (3), (4) and (5) all converge to $\gamma a_x^1 \sigma^2$. Term (2) is an average of $\ep[i]^3$, which is mean zero; consequently, it converges to 0.

This completes the proof.

@@

>>>>

\begin{figure*}[h]
	\centering
	\begin{subfigure}[h]{0.33\linewidth}
		\centering
		\includegraphics[width=.8\linewidth]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\linewidth}
		\centering
		\includegraphics[width=.8\linewidth]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\linewidth}
		\centering
		\includegraphics[width=.8\linewidth]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of micrographs of size $250\times 250$ with additive white Gaussian noise of variance $\sigma^2$ for increasing values of $\sigma$. Each micrograph contains the same four occurrences of a $50 \times 50$ image of Einstein. In panel (c), the noise level is such that it is very challenging to locate the occurrences of the planted image. In fact, it can be shown that at low SNR, reliable detection of individual image occurrences is impossible, even if the true image is known. By analogy to cryo-EM, this depicts a scenario where particle picking cannot be done.}
	\vspace{-12pt} 	
\end{figure*}

\begin{figure*}[h]
	\centering
	\begin{subfigure}[h]{0.24\linewidth}
		\centering
		\includegraphics[width=1.1\linewidth]{Einstien_progress_1}
		\caption{$2\times 10^2$ micrographs}
	\end{subfigure} \hfill
	\begin{subfigure}[h]{0.24\linewidth}
		\centering
		\includegraphics[width=1.1\linewidth]{Einstien_progress_4}
		\caption{$2\times 10^3$ micrographs}
	\end{subfigure} \hfill
	\begin{subfigure}[h]{0.24\linewidth}
		\centering
		\includegraphics[width=1.1\linewidth]{Einstien_progress_40}
		\caption{$2\times 10^4$ micrographs}
	\end{subfigure} \hfill
	\begin{subfigure}[h]{0.24\linewidth}
		\centering
		\includegraphics[width=1.1\linewidth]{Einstien_progress_400}
		\caption{$2\times 10^5$ micrographs}
	\end{subfigure}
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized randomly. Estimates are obtained from $2\times 10^2,2\times 10^3,2\times 10^4,2\times 10^5$ micrographs (growing across panels), each containing $700$ image occurrences on average.}	
	\vspace{-12pt} 
\end{figure*}

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical SNR. This fact has been recognized early on by the cryo-EM community. In particular, in an influential paper from 1995, Richard Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{myquote}{0.1in}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{myquote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight (below $\sim$50 kDa). 
Joachim Frank voices a similar observation in his 2017 Nobel prize lecture: ``\emph{Using the ribosome as an example, it became clear from the formula we obtained that the single-particle approach to structure research was indeed feasible for molecules of sufficient size: Particle Size > 3/[Contrast$^2$ $\times$ Resolution (as length) $\times$ Critical Electron Dose]}''~\cite{frank2018single}. 
As these two leaders of the cryo-EM community point out, it is impossible to reconstruct small molecules by any of the existing computational pipelines for single particle analysis in cryo--EM, because the particles themselves cannot be picked from the micrographs. Yet, imaging such molecules is important for a number of applications~\cite{scapin2018cryo}

@@

We observe $X$, and our task is to estimate $\eta$. How reliably can this be done? If $q \geq 1/2$, the constant estimator $\hat{\eta} = 0$ is correct with probability $q$; likewise, if $q \leq 1/2$, the constant estimator $\hat{\eta} = 1$ is correct with probability $1-q$. The question is, can we do better than this?  We prove that, as $\sigma \to \infty$, the answer is no. The result is proved in Appendix~\ref{sec:proof_two_gauss}.
\begin{proposition} \label{prop:two_gauss}
	For any deterministic estimator $\hat{\eta }$ of $\eta$,
	\begin{align*}
	\lim_{\sigma \to \infty} \P[ \hat{\eta} = \eta  ] & \leq \max(q, 1-q);
	\end{align*}
	that is: as the SNR deteriorates, the probability of success is no better than random chance.
\end{proposition}

@@

%From Propositions~\ref{prop:uniqueness} and~\ref{prop:gamma_sigma} we   deduce the following:
%\begin{corollary}
%	%Let $K=1$ and $\sigma > 0$ be fixed.
%	In the limit of $N\to \infty$ and under the separation condition~\eqref{eq:spacing}, the signal $x$, the ratio $\gamma$, and the noise level $\sigma$ are determined from the first three autocorrelation functions of $y$ if either the signal $x$ is generic or $x$ has nonzero mean  and $\gamma > \frac{1}{4}$.
%\end{corollary}

%This can be added to an arxiv version of the paper	
From Propositions~\ref{prop:uniqueness} and~\ref{prop:gamma_sigma} we   deduce the following:
\begin{corollary}
	%Let $K=1$ and $\sigma > 0$ be fixed.
	 In the limit of $N\to \infty$ and under the separation condition~\eqref{eq:spacing}, the signal $x$, the ratio $\gamma$, and the noise level $\sigma$ are determined from the first three autocorrelation functions of $y$ if:
	\begin{itemize}
		\item Either the signal $x$ is generic; or
		\item   $x$ has nonzero mean  and $\gamma > \frac{1}{4}$.
	\end{itemize}
\end{corollary}
%
%\TODO{Similar results hold for signals in higher dimensions.---need to keep this sentence in red until the very end to make sure it's correct.}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{Einstein_recovery_error_combined}
	\caption{\label{fig:error_per_micro}Relative error curves for the experiment of Figure~\ref{fig:Einst_example}.}
\end{figure}

@@

@@

{\bf Case 4:} $0=\ell_2 = \ell_1$. The same argument as for term (4) in Case 2 shows that terms (3), (4) and (5) all converge to $\gamma a_x^1 \sigma^2$. Term (2) is an average of $\ep[i]^3$, which is mean zero; consequently, it converges to 0.
This completes the proof.
